Amplitude metrics for cellular \\ circadian bioluminescence reporters

### 2014-08-29 main_text.tex
stephanie's changes, better explanations of the numerical methods in fig. 5

The second contribution to the perturbed population trajectory comes from deviations from limit cycle oscillations in each cell.
We calculate the population-level effect of these deviations by averaging over the deviations that occur at each phase.
We define the deviation trajectory $\delta x(\theta, \tilde{t})$ for each phase as the distance between the perturbed trajectory and the phase-adjusted reference:
\begin{align}
  \delta x(\theta_0, \tilde{t}) &\coloneqq x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0 + \Delta \theta) \\
  \therefore\; \lim_{\tilde{t} \to \infty} \delta x(\theta, \tilde{t}) &= 0
\end{align}
Since the perturbed trajectory ultimately converges with the phase-adjusted reference, deviations will converge to zero.
Since the phase change, $\Delta\theta$, associated with a perturbation at each phase is likely not known prior to calculating the perturbed trajectory, it is difficult to tabulate deviation trajectories associated with each final phase, $\theta_0 + \Delta\theta$.
It is therefore more straightforward to find the average effect of single-cell perturbations at the population level by weighting the deviations by the phase density function prior to perturbation.
The population-level response to a perturbation, $\hat{x}(\tilde{t})$, is therefore defined as:
\begin{equation}
  \hat{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta)\hat{p}(\theta, \tilde{t}) + \delta x(\theta, \tilde{t})p(\theta, \tilde{t}) \; d\theta
\end{equation}
Here the first term is equivalent to $\hat{x}_{ss}(\tilde{t})$, the steady-state perturbed trajectory, while the second term decays to zero as individual oscillators return to their steady-state amplitude.
The accuracy of the continuous phase-diffusion model was tested by explicitly simulating a perturbation to a population of 225 uncoupled oscillators using a stochastic simulation algorithm \cite{Gillespie1977, Sanft2011a}.
Good agreement between the stochastic population and continuous approximation was shown in phase shift, single-cell level amplitude change, and alteration of population synchrony (Fig.~S2, Movies S1 \& S2).
These results indicate the use of a continuous probability function is justified in the case of cultured cellular reporter systems, which may contain up to $10^5$ individual cells per culture \cite{Welsh2004}.
In addition, the method was verified by similar simulations using the Oregonator \cite{Field1974}, a model containing only mass-action terms, and a detailed mechanistic model of circadian rhythms \cite{Hirota2012} at a variety of phase timings (Fig.~S3).
The continuous approximation is therefore able to capture the population-level dynamics of a wide variety of limit cycle models and perturbations.

@@

Data on the response of melanopsin-responsive and control NIH3T3 mouse fibroblasts cells is shown in Fig.~5B \cite{Ukai2007}.
The experimental data is de-noised using a discrete wavelet transform \cite{Leise2011}.
In the experiment, a light pulse is given to desynchronize a colony of cells, resulting in drastically reduced amplitudes.
A second (longer) light pulse subsequently re-synchronizes the population, resulting in increased amplitudes.
To capture this phenomenon {\itshape in silico}, we used a recent model of the core circadian feedback circuit \cite{Hirota2012}.
We assumed an initial phase population with $\sigma = 0.1$ and calculated a phase diffusivity parameter $d = 0.104$ from the experimental control trajectory.

@@

The simulated control and perturbed trajectories, Fig.~5B, closely match the experimental results, in which the model captures both the phase shifts and amplitude modulation of the light pulses.
The phase probability density function for the light-sensitive model trajectory is shown in Fig.~5C for several representative time points.
Changes to the synchronization of the population from each perturbation are readily apparent, with both perturbations inducing a bimodal distribution in the phase density.
While experimental data on individual cell phases was unavailable for this data set, bimodal distributions in SCN neuron firing following a phase shift have previously been seen experimentally \cite{Rohling2011}.
As higher-frequency features, these bimodal distributions dissipate as the phases diffuse.

>>>>

The second contribution to the perturbed population trajectory comes from deviations from limit cycle oscillations in each cell.
We calculate the population-level effect of these deviations by averaging over the deviations that occur at each phase.
We define the deviation trajectory $\delta x(\theta, \tilde{t})$ for each phase as the distance between the perturbed trajectory and the phase-adjusted reference:
\begin{align}
  \delta x(\theta_0, \tilde{t}) &\coloneqq x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0 + \Delta \theta) \\
  \therefore\; \lim_{\tilde{t} \to \infty} \delta x(\theta, \tilde{t}) &= 0
\end{align}
Since the perturbed trajectory ultimately converges with the phase-adjusted reference, deviations will converge to zero.
Since the phase change, $\Delta\theta$, associated with a perturbation at each phase is likely not known prior to calculating the perturbed trajectory, it is difficult to tabulate deviation trajectories associated with each final phase, $\theta_0 + \Delta\theta$.
It is therefore more straightforward to find the average effect of single-cell perturbations at the population level by weighting the deviations by the phase density function prior to perturbation.
The population-level response to a perturbation, $\hat{x}(\tilde{t})$, is therefore defined as:
\begin{equation}
  \hat{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta)\hat{p}(\theta, \tilde{t}) + \delta x(\theta, \tilde{t})p(\theta, \tilde{t}) \; d\theta
  \label{eq:xhat}
\end{equation}
Here the first term is equivalent to $\hat{x}_{ss}(\tilde{t})$, the steady-state perturbed trajectory, while the second term decays to zero as individual oscillators return to their steady-state amplitude.
The accuracy of the continuous phase-diffusion model was tested by explicitly simulating a perturbation to a population of 225 uncoupled oscillators using a stochastic simulation algorithm \cite{Gillespie1977, Sanft2011a}.
Good agreement between the stochastic population and continuous approximation was shown in phase shift, single-cell level amplitude change, and alteration of population synchrony (Fig.~S2, Movies S1 \& S2).
These results indicate the use of a continuous probability function is justified in the case of cultured cellular reporter systems, which may contain up to $10^5$ individual cells per culture \cite{Welsh2004}.
In addition, the method was verified by similar simulations using the Oregonator \cite{Field1974}, a model containing only mass-action terms, and a detailed mechanistic model of circadian rhythms \cite{Hirota2012} at a variety of phase timings (Fig.~S3).
The continuous approximation is therefore able to capture the population-level dynamics of a wide variety of limit cycle models and perturbations.

@@

Data on the response of melanopsin-responsive and control NIH3T3 mouse fibroblasts cells is shown in Fig.~5B \cite{Ukai2007}.
The experimental data is de-noised using a discrete wavelet transform \cite{Leise2011}.
In the experiment, a light pulse is given to desynchronize a colony of cells, resulting in drastically reduced amplitudes.
A second (longer) light pulse subsequently re-synchronizes the population, resulting in increased amplitudes.
To capture this phenomenon {\itshape in silico}, we used a recent model of the core circadian feedback circuit \cite{Hirota2012}.
In order to find parameters for the phase density function, we assumed an
initial phase population with $\sigma = 0.1$ and calculated a phase diffusivity
parameter $d = 0.104$ from the experimental control trajectory.

@@

Population-level trajectories were found by solving for the resulting limit cycle trajectories at every phase, and subsequently finding the weighted average of these trajectories according the phase density (as described in Eqs.~\ref{eq:xbar} and \ref{eq:xhat}).
The simulated control and perturbed trajectories, Fig.~5B, closely match the experimental results, in which the model captures both the phase shifts and amplitude modulation of the light pulses.
The phase probability density function for the light-sensitive model trajectory is shown in Fig.~5C for several representative time points.
Changes to the synchronization of the population from each perturbation are readily apparent, with both perturbations inducing a bimodal distribution in the phase density.
While experimental data on individual cell phases was unavailable for this data set, bimodal distributions in SCN neuron firing following a phase shift have previously been seen experimentally \cite{Rohling2011}.
As higher-frequency features, these bimodal distributions dissipate as the phases diffuse.
